Sampling Strategies in Decisions from Experience

### 2021-12-23 manuscript.Rmd
Finished Draft Introduction

The mind is a cognitive system that operates on inputs from an environment or memory.
Such inputs may be treated as a set $I$ and its elements, the latter being distinguishable from each other on the basis of the *properties* they possess [cf. @jostMathematicalConcepts2015].
Often, however, the properties that the mind uses to control the output are not readily described, but must be inferred from incomplete or noisy information. 
Since how the inputs are represented determines the information that is made explicit and the costs at which certain operations can be carried out on that information, the feasibility and ease of these inferences depends greatly on the choice of *representation* [@marrVisionComputationalInvestigation1982; see also @griffithsProbabilisticModelsCognition2010; @kempStructuredStatisticalModels2009].
Hence, to explain how the mind turns certain inputs into meaningful outputs, one needs to understand the properties that control the output, the representation that is used to describe these properties, and the costs of making the descriptions in that representation, including their susceptibility to distortions from uncertainty or noise.

In what follows, the discussion will turn to the domain of decision making and elaborate on the concepts touched upon thus far, certainly only a subset of the factors involved in the decision making process, yet central enough to illustrate one rather important mechanism:
Given that the properties of choice alternatives serve as *selection criteria* and the mind is indeed sensitive to variations in their values, the more accurately these properties are described, the more systematic the choices should become.
Importantly, this principle is not restricted to any one specific property, but rather does the choice pattern reflect the bias that the designation of a property as a selection criterion imposes on the assessment of choice alternatives.

@@

Decision making is an information-processing problem whose solution requires a selection among choice alternatives. 
The basic structure of the problem may be modeled by the inclusion map $\iota : I' \mapsto I$, where $\iota(i) = i$ for all $i \in I'$, indicating that decision making is a *process* that starts from an input set $I$, including all choice alternatives, and stops at a proper subset $I' \subset I$, including only the chosen alternatives. 
Given that the selection process is systematic---as opposed to random---the chosen alternatives in $I'$ should possess the same property or combination of several properties which the other alternatives in the complementary subset $I \setminus I'$ do not possess.
We may label the properties that serve to distinguish between the chosen and non-chosen alternatives as *selection criteria*.

@@

<!-- In the same vein, there is a call for *representational pluralism* in cognition research [@griffithsProbabilisticModelsCognition2010] -->

@@

# Capturing Differences in Sampling and Decision Strategies

@@

# Predicting Choice Behavior

@@

# Method

## Generating Model

@@

## Describing Model: Cumulative Prospect Theory

@@

# Results

## Plausibility Check: Relationship between Switching Probability and Trial Length

@@

## Probability Weighting Function

@@

## Value Function

@@

# Discussion

# Conclusion

@@

# Software

>>>>

The mind is a cognitive system that operates on inputs from an environment or memory.
Such inputs may be treated as a set $I$ and its elements, the latter being distinguishable from each other on the basis of the *properties* they possess [cf. @jostMathematicalConcepts2015].
Often, however, the properties used by the mind to control the outputs are not readily described, but proper descriptions must be inferred from incomplete or noisy information. 
Since how the inputs are represented determines the information that is made explicit and the costs at which certain operations can be carried out on that information, the feasibility and ease of these inferences depends greatly on the choice of *representation* [@marrVisionComputationalInvestigation1982; see also @griffithsProbabilisticModelsCognition2010; @kempStructuredStatisticalModels2009].
Hence, to explain how the mind turns certain inputs into meaningful outputs, one needs to understand the properties that control the output, the representation that is used to describe these properties, and the costs of inferring the descriptions in that representation, including their susceptibility to distortions from uncertainty or noise.

In what follows, the discussion will turn to the domain of decision making and elaborate on the concepts touched upon thus far, certainly only a subset of the factors involved in the decision making process, yet central enough to illustrate one rather important mechanism:
Given that the properties of choice alternatives serve as *selection criteria* and the mind is indeed sensitive to variations in their values, the more accurately the descriptions of these properties are inferred, the more systematic the choices should become.
Importantly, this principle is not restricted to any one specific property, but rather does the choice pattern result from the constraints that the designation of a property as a selection criterion imposes on the assessment of choice alternatives.

As the discussion proceeds and for the sake of example, I argue that a choice pattern robustly emerging in so-called *decisions from experience* (DfE), i.e., the apparent underweighting of small probability events, may be caused by the assessment constraints associated with a round-wise decision rule.
This particular argument has already been made by @hillsInformationSearchDecisions2010, however, decision rules, as I try to explain below, are for several reasons better formulated as selection criteria whose descriptions are rarely restricted to one representation.
Considering *sampling strategies* as representations used to describe the properties of interest in DfE, I illustrate how the inferred descriptions of these properties indeed become more accurate with increasing sample sizes. 
Yet, caused by the selection criterion corresponding to the round-wise decision rule, the underweighting pattern is a priori predicted to become more pronounced in such a case. 
This prediction is different from the relatively undisputed explanation that underweighting is caused by the reliance on small samples and the skewness of the binomial distribution, indicating that the effects of sample size on underweighting are conditional on the selection criterion. 
I will elaborate on this matter as I proceed, however, to be rather clear from the outset, what the mind does in neither case is performing an underweighting computation in its literal sense, but it processes the inputs in a way that the outputs approximate the solutions of what can be considered probability weighting on an abstract computational level.

The theoretical argument is substantiated by a simulation study [cf. @guestHowComputationalModeling2021]. 
For this purpose, the sampling and decision processes are expressed mathematically and implemented in a computational model.
Choice data from DfE is simulated and modeled using cumulative prospect theory (CPT), a data model commonly used for describing decisions under risk and uncertainty [@tverskyAdvancesProspectTheory1992]. 
All simulations and analyses are reproducible---materials and instructions can be found on the accompanying GitHub repository.

# Heading Section 1

@@

Decision making is an information-processing problem whose solution requires a selection among choice alternatives. 
The basic structure of the problem may be modeled by the inclusion map $\iota : I' \mapsto I$, where $\iota(i) = i$ for all $i \in I'$, indicating that decision making is a *process* that starts from an input set $I$, including all choice alternatives, and stops at a proper subset $I' \subset I$, including only the chosen alternatives. 
Given that the selection process is systematic---as opposed to random---the chosen alternatives in $I'$ should possess the same property or combination of several properties which the other alternatives in the complementary subset $I \setminus I'$ do not possess.
We may label the properties that serve to distinguish between the chosen and non-chosen alternatives as selection criteria.

@@

# Heading Section 2

@@

## Subheading: Capturing Differences in Sampling and Decision Strategies

@@

## Subheading: Predicting Choice Behavior

@@

# Simulation Study

@@

## Results

@@

### Plausibility Check: Relationship between Switching Probability and Trial Length

@@

### Probability Weighting Function

@@

### Value Function

@@

# Discussion and Conclusion

@@
